---
# Reproducible Research: Peer assessment 1 
### *Syada Chowdhury* 
### *19 APRIL 2015* 
---
# Change working directory 
setwd("C:/Users/Syada/Desktop/Coursera/Reproducible") 

#First, we set echo equal a TRUE and results equal a 'hold' as global options for this document. 

```{r}
echo=TRUE
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')

````


#Loading and preprocessing the data
#Loading the required data.

```{r}
data <- read.csv("activity.csv", stringsAsFactors = FALSE)

```

#Process/transform the data into a format by converting the date to a valid R date object

```{r}
data$date <- as.Date(data$date)
```

#Now, let us check the data using str() method:
```{r}
str(data)
```
#What is mean total number of steps taken per day?
#Now here we ignore the missing values(a valid assumption).
#Calculate the total number of steps taken per day
```{r}
steps_per_day <- aggregate(steps ~ date, data, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)
```

#Creating a histogram of the total number of steps taken per day, plotted with appropriate bin interval
```{r}
ggplot(steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "green", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 

```

#Calculate and report the mean and median of the total number of steps taken per day

```{r}
steps_mean   <- mean(steps_per_day$steps, na.rm=TRUE)
steps_median <- median(steps_per_day$steps, na.rm=TRUE)

````
#What is the average daily activity pattern?

#1.Making a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)


```{r}
averageStepsByInterval <- tapply(data$steps, data$interval, mean, na.rm = TRUE)
plot(names(averageStepsByInterval), averageStepsByInterval, type = "l", xlab = "5-minute Interval", 
    ylab = "Average number of steps taken", main = "Avg number of steps across 5-minute interval")

```


# Calculate Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps

```{r}
# Find the index and the maximum value
maxValues <- which.max(averageStepsByInterval)[1]
maxInterval <- names(which.max(averageStepsByInterval))
maxValue <- which.max(averageStepsByInterval)[[1]]
# Convert the interval from string to integer value
maxInterval <- strtoi(maxInterval, 10L)

maxInterval
## [1] 835
maxValue
## [1] 104
```

#Imputing missing values
#1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
missingValues <- is.na(data$steps)
table(missingValues)[[2]]
## [1] 2304

```


# 2.Strategy for filling in all missing values in the dataset ,by using the mean for the 5-minute interval.

```{r}
meanStepsByInterval <- tapply(data$steps, data$interval, mean, na.rm = TRUE)
naRows <- data[is.na(data), ]
for (i in 1:nrow(naRows)) {
    naRows[i, 1] <- meanStepsByInterval[[toString(naRows[i, 3])]]
}


```

#Creating a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
filledInData <- data
filledInData[is.na(data), 1] <- naRows[, 1]
str(filledInData)

#'data.frame':  17568 obs. of  3 variables:
# $ steps   : num  1.717 0.3396 0.1321 0.1509 0.0755 ...
# $ date    : Date, format: "2012-10-01" "2012-10-01" ...
# $ interval: int  0 5 10 15 20 25 30 35 40 45 ...

```

#Checking that there are any missing values remaining or not

```{r}
sum(is.na(filledInData$steps))
## [1] 0
```

#Zero output shows that there are NO MISSING VALUES.


#4.Now let us plot a histogram of the daily total number of steps taken, plotted with a bin interval of 1000 steps, after filling missing values.

```{r}
fill_steps_per_day <- aggregate(steps ~ date, filledInData, sum)
colnames(fill_steps_per_day) <- c("date","steps")

##plotting the histogram
ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
```

#Calculate and report the mean and median total number of steps taken per day.

```{r}
steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
```
#The mean is 10766.189 and median is 10766.189.

#Do these values differ from the estimates from the first part of the assignment?
```{r}

##Yes, these values do differ slightly.
### Before filling the data
# 	1.Mean : 10766.189
# 	2.Median: 10765
###After filling the data
# 	1.Mean : 10766.189
# 	2.Median: 10766.189
###We see that the values after filling the data mean and median are equal.
```

##What is the impact of imputing missing data on the estimates of the total daily number of steps?

## As We can see, comparing with the calculations done in the first section of this document, we observe that while the mean value remains unchanged, the median value has shifted and virtual matches to the mean. 
##Since our data has shown a t-student distribution (see both histograms), it seems that the impact of imputing missing values has increase our peak, but it's not affect negatively our predictions. 


##Are there differences in activity patterns between weekdays and weekends?

```{r}
days <- weekdays(filledInData$date)
weekend <- (days == "Saturday" | days == "Sunday")
dayfactor <- factor(weekend, labels = list("weekday", "weekend"))
filledInData$daytype <- dayfactor

groupedData <- aggregate(filledInData$steps, list(DayType = filledInData$daytype, 
    Interval = filledInData$interval), mean)
library(lattice)
xyplot(groupedData$x ~ groupedData$Interval | groupedData$DayType, layout = c(1, 
    2), xlab = "Interval", ylab = "Number of Steps", type = "l")
```

##From the above graph we can see that activity on the weekday has the greatest peak from all steps intervals apparently weekend activities has more peaks over a hundred than weekday. This might be due to the fact that activities on weekdays mostly follow a work related routine, where we find some more intensity activity in little a free time that the employ can made some sport. On the other hand, at weekend a better distribution of effort along the time is showing in a good way.


